---
title: "Mixed Effects Model"
format: html
editor: visual
bibliography: references.bib
---

## Resources

**Mixed Effects Models and Extensions in Ecology with R**: https://link.springer.com/book/10.1007/978-0-387-87458-6

**Linear Mixed Effects Modelling:** <https://wiki.qcbs.ca/r_workshop6>

**Check for temporal autocorrelation in the model residuals:** <https://bbolker.github.io/mixedmodels-misc/notes/corr_braindump.html>

**With and without temporal autocorrelation:** <https://bbolker.github.io/mixedmodels-misc/notes/corr_braindump.html>

**Mixed effects model tutorial:** <https://ourcodingclub.github.io/tutorials/mixed-models/index.html#what>

chrome-[extension://efaidnbmnnnibpcajpcglclefindmkaj/https://bodowinter.com/tutorial/bw_LME_tutorial.pdf](extension://efaidnbmnnnibpcajpcglclefindmkaj/https://bodowinter.com/tutorial/bw_LME_tutorial.pdf)

```{r}
# Sample Code
lme1 <- lme(logFCH4_gC~GPP_PI_F_NT_gC+logTP_interp+TA+logNO3_NO2_N_interp+WTD+logSO4_pred_interp,random=~1|site,data=data.daily.model.norm,method = mth)
lme2 <- lme(logFCH4_gC~GPP_PI_F_NT_gC+logTP_interp+TA+logNO3_NO2_N_interp+WTD+logSO4_pred_interp,random=~1|site,data=data.daily.model.norm,method = mth,correlation=corAR1()) # Best - same as lme3
lme3 <- lme(logFCH4_gC~GPP_PI_F_NT_gC+logTP_interp+TA+logNO3_NO2_N_interp+WTD+logSO4_pred_interp,random=~1|site,data=data.daily.model.norm,method = mth,correlation=corAR1(form = ~ 1 | site))
AICc<-c(AICc(lme1), AICc(lme2),AICc(lme3))
# Put values into one table for easy comparison
Model<-c("LME1", "LME2","LME3")
AICtable<-data.frame(Model=Model, AICc=AICc)
AICtable
# Refit final model
m.final <- lme(logFCH4_gC~GPP_PI_F_NT_gC+logTP_interp+TA+logNO3_NO2_N_interp+WTD+logSO4_pred_interp,random=~1|site,data=data.daily.model.norm,method = "REML",correlation=corAR1())
summary(m.final)
# 6) Model validation
# Checking model assumptions
# a) Look at independence: plot fitted values vs residuals
E1 <- residuals(m.final) # Select final model
F1<-fitted(m.final)
plot(x = F1,
     y = E1,
     xlab = "Fitted Values",
     ylab = "Normalized residuals")
abline(h = 0, lty = 2)
data.daily.model.norm$F1 <- F1
data.daily.model.norm$E1 <- E1
```

## Similar Research that Use the Mixed Effects Model

-   <https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2017JG004227>

-   <https://onlinelibrary.wiley.com/doi/full/10.1111/gcb.13612>

-   <https://www.sciencedirect.com/science/article/pii/S0038071712001927>

-   23/11:

-   <https://agupubs.onlinelibrary.wiley.com/doi/10.1029/2021JG006328>

-   <https://agupubs.onlinelibrary.wiley.com/action/downloadSupplement?doi=10.1029%2F2021JG006328&file=2021JG006328-sup-0002-Table+SI-S01.pdf>

## Overview of the Model

1.1 Mixed Effects Model can solve the problem of:

-   Messy of the data: relationships between variables of interest might differ depending on grouping factors like species

-   Sample sizes are low making it difficult to fit models that require many parameters to be estimated

1.2 Two types of effects in the model:

1\) Fixed effect: when a variable has a fixed effect, data is usually gathered from all its possible levels. The person doing the analyses is also interested in making conclusions about the levels for which the data was gathered.

2\) Random effect: variables with a random effect are also called random factors, as they are only categorical variables (not continuous variables). A random effect is observed when the data only includes a random sample of the factor's many possible levels, which are all of interest. They usually are grouping factors for which you want to control the effect in your model, but are not interested in their specific effect on the response variable.

1.3 How do LMM work?

## Notes from Linear mixed Effects Model Workshop (<https://wiki.qcbs.ca/r_workshop6>)

```{r}
# Remove prior commands in R
rm(list=ls()) 
 
# Place all workshop material in one folder on your computer
# Run the following line of code and use the browsing window to choose the QCBS_W6_Data.csv 
# file in the folder that contains the workshop material
file.choose()
 
# Set the working directory to the folder which contains the lab material by copy and pasting
# all but the R file name from the output of the file.choose() command into the set working 
# directory command. 
 
# For example paste "/Users/ziegljac/Documents/QCBS_R/" -> include the quotations
# NOT "/Users/ziegljac/Documents/QCBS_R/Get_Data_Func.R" -> include the quotations
 
# Load useful libraries and data
# If you have never loaded these libraries before you will have to use 
# The install.packages() function before the library() function
library(ggplot2)
library(lme4)
library(arm)
library(AICcmodavg)

setwd("/Users/zoe/Literature")
data <- read.csv('fishdata.csv')
```

```{r}
# Step 1: Always check the data structure
str(data)
# In R, the table function is used to produce a frequency table of categorical data.
table(data$Lake)
table(data$Fish_Species)
# Look at distribution of continuous variables
# Major skews can lead to problems with model homogeneity down the road. So if necessary, make transformations. In this case, the data seems OK.
hist(data$Fish_Length)
hist(data$Trophic_Pos)
# Check for collinearity between variables: It is important not to include 2 collinear variables in the same model, otherwise their effects on the response variable will be confounded, i.e. the model can't tell which of the collinear variable is responsible for the variation in the response variable. 
plot(data)
cor(data$Fish_Length, data$Var2)
```

```{r}
# Step 2: Consider the scale of your data
# If two variables in the same model have different ranges of scale, the criteria the mixed models use to come up with parameter estimates are likely to return 'convergence errors'. Z-correction standardizes your variables and adjusts for this scaling problem by placing all your variables on the same scale even if they were originally in different units:
# z = (x-mean(x))/sd(x)
# Z-correct Length
data$Z_Length<-(data$Fish_Length-mean(data$Fish_Length))/sd(data$Fish_Length)
# Z-correct Trophic Position
data$Z_TP<-(data$Trophic_Pos-mean(data$Trophic_Pos))/sd(data$Trophic_Pos)
```

\# To know if a mixed model is necessary for your data set you must establish if it is actually important to account for variation in the factors that might be affecting the relationship that you're interested in, Lake and Species in this case. We can do this by: - Creating a linear model without factors - Calculating the residuals of this linear model - Plotting the residuals of the linear model against the factors of interest

```{r}
# Find out if it is important to account for variation in "random effects"
# by comparing the residuals of a linear model without the random effects with 
# the potential random effects
lm.test<-lm(Z_TP~Z_Length, data=data)
lm.test.resid<-rstandard(lm.test)
# Species Effect
plot(lm.test.resid~ data$Fish_Species, xlab = "Species", ylab="Standardized residuals")
abline(0,0, lty=2)
# Lake Effect
plot(lm.test.resid~ data$Lake, xlab = "Lake", ylab="Standardized residuals")
abline(0,0, lty=2)
```

## Literature Review

**Wildfire overrides hydrological controls on boreal peatland methane emissions** [@davidson2019]

From 2017-2018, measured CH4 emissions, environmental variables and CH4 production potential in 2 different microform types (hummocks & hollows) across a peat burn severity gradient (unburned-UB; moderately burned-MB and severely burned - SB)

Methods:

-   To evaluate the effect of burn severity on CH4 flux, a linear mixed effects model (LMM) was used with burn severity, microform, the two-way interactions between these, and year as fixed factors, with collar ID as a random effect to take into account repeated measures.

-   Another LMM was used to evaluate the environmental controls on CH4 flux, with the effect of burn severity, WT, ST at 30cm depth and the two-way interactions between each included as fixed effects.

-   If significant factors were found, Tukey pairwise comparisons were completed using the lsmeans package.

**Mixed Effect Models and Extensions in Ecology with R**

1.  Linear regression models
2.  Chapter 4 -\> deal with heterogeneity in linear regression
3.  Chapter 5 -\> Random Effects for nested data
4.  Chapters 6 and 7 -\> temporal and spatial correlation

Chapter 3 - Additive Modelling, generalized additive modeling (GAM)

Eg: Pelagic bioluminescence along a depth gradient in the northeast Atlantic Ocean

The additive model fits a smoothing curve through the data.

Selecting Among Alternative Models

1.  Meeting assumptions: Model must meet the assumption of a linear relationship to be come a candidate linear regression model
2.  Comparing the fit statistics. Select the higher R2 and lower RMSE.
3.  Reject any models where the regression is not significant.
4.  A simpler model is preferred

## Resources

The seedling removal project - growing season:

-   Refer to the Mixed Effects Models and Extensions with Ecology in R, check example 2.3.8, for heterogeneity

Notes from <https://ourcodingclub.github.io/tutorials/mixed-models/index.html#what>:

There are two types of random effects:

**1) Crossed random effects**: multiple observations on one examine sample, such as chamber flux measurements (multiple flux observations has been collected on a chamber across time)

**2) Nested random effects:** there\'s often an element of scale, or sampling stratification in there (you have 6 collars in a study site)
